The evolution of gene expression in the term placenta of viviparous mammals

### 2015-08-28 manuscript/manuscript.Rnw
add more missing references to the manuscript

One symptom
of this maternal-fetal conflict is its potential affect on human
health (Hollegaard et al., 2013). In addition to having an invasive
hemochorial placenta, humans possess a deeply invasive trophoblast
layer that can penetrate the myometrium of the uterus (Carter and
Pijnenborg, 2011). Defects in placental growth are known to be
associated with the devastating condition of preeclampsia, which can
cause hypertension, proteinuria, and has the potential to lead to
death of the mother and/or infant (citation). Although cases have been
documented in chimpanzees and gorillas (citation), preeclampsia has a
far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States (citation).
Chimpanzees and gorillas also have deeply invasive trophoblasts,
although the progression of invasion in great apes may be slower than
humans (Carter and Pijnenborg, 2011); however, large-scale expression
studies of great ape placentas have not been conducted. The increasing
affordability of next-generation sequencing technology has enabled the
large-scale study of specific tissues from a wide array of species to
address broader phylogenetic questions in the evolution of gene
expression (Brawand et al., 2011). Here we use Illumina RNA-Seq
technology to sequence and quantify the mRNA environment of six
mammalian placentas at term, including that of the human, bonobo (the
sister species to the chimpanzee), dog, cow, African elephant, and
opossum, representing marsupials as well as three major branches of
eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation (Table 1). We use this data, along with existing RNA-Seq
data from mouse (Wang et al., 2011), to address the major differences
between expression in marsupial and eutherian mammals at term,
identify the evolution of expression in important placental genes such
as hormones and galectins, identify novel lineage-specific expression
patterns, and quantify human specific differences in gene expression
that may be related to obstetrical disorders such as preeclampsia.

@@

Fetal tissue was collected from the placenta of six mammalian species
(Homo sapiens [human], Pan paniscus [bonobo], Canis familiaris
[domestic dog], Bos taurus [domestic cow], Loxodonta africana [African
savanna elephant], and Monodelphis domestica [gray short-tailed
opossum]). Information on specimen ID, place and time of collection,
and collector is shown for each sample in Supplemental Table 1. In the
case of Pan, Bos, and Loxodonta, the placenta was collected at term
upon expulsion from the mother following delivery. Placental tissue
from Homo sapiens was collected upon delivery via a caesarian section.
Near-term placental tissue from Canis familiaris was collected upon
the death of the mother from a car accident. Placental tissue from
Monodelphis domestica was sampled at fetal stage 32, roughly one day
prior to birth. Only the fetal portion of the placenta was sampled at
the maternal-fetal interface (i.e. villous portion for Homo, Pan,
Canis, and Loxodonta; cotyledon for Bos; diffuse placenta for
Monodelphis). All tissues were collected and stored in RNAlater (Life
Technologies, Carlsbad, CA) until RNA isolation could be performed.
Tissues were homogenized in Trizol, and total RNA was isolated
following the manufacturer’s instructions. RNA was further purified
using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction with the
RNase-Free DNase Set (Qiagen, Valencia, CA), following the
manufacturer’s protocol to remove any contaminating DNA from the
sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
Clara, CA) to confirm concentration and quality of the isolation. RNA
quality was assessed using A260/A280 and A260/A230 absorbance values.

@@

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and standard Illumina protocol. Paired-end sequencing was
performed using the Illumina Genome Analyzer II with an insert size of
230 base pairs (bp). Bos, Loxodonta, and Monodelphis were sequenced
with read lengths of 75 bp; Canis, Homo, and Pan were sequenced with
read lengths of 150 bp.

@@

The quality of reads was evaluated using FASTQC v. 0.9.2
(http://www.bioinformatics.babraham.ac.uk/projects/fastqc/). Following
this, reads were trimmed based on quality in Galaxy using the FASTQ
Trimmer v. 1.0.0 (Blankenberg et al., 2010). All transcriptomes were
then aligned to reference genomes and assembled using the Tuxedo suite
of next-generation sequencing analysis tools. Annotated reference
genomes were downloaded from Ensembl v. 64 (Flicek et al., 2012),
including Homo sapiens GRCh37, Pan troglodytes CHIMP2.1 (6X coverage),
Canis familiaris BROADD2 (7.6X), Bos taurus UMD3.1 (9.5X), Loxodonta
africana loxAfr3 (7X), and Monodelphis domestica BROADO5 (6.8X). The
genome of the closely-related Pan troglodytes was used to map the
transcriptome of Pan paniscus; the separation between the two
recognized Pan species is generally perceived to be ~1 Mya (Prufer et
al., 2012). Reads were aligned to reference genomes using Bowtie v.
0.12.7 (Langmead et al., 2009) and TopHat v. 1.3.3 (Trapnell et al.,
2009). Cufflinks v. 1.2.1 was then used to assemble transcripts and
quantify expression values using fragments per kilobase of exon per
million mapped reads (FPKM) (Trapnell et al., 2010). RNA-Seq metrics
were computed using Picard (http://picard.sourceforge.net/). For
comparison, we also downloaded the placenta Illumina RNA-Seq
transcriptome of a mouse from (Wang et al., 2011). Raw reads were
mapped as above using the Ensembl v. 64 annotation of the Mus musculus
(house mouse) genome (NCBI m37).

@@

The human placenta is distinctive from many other primates in that it
possesses a deeply invasive trophoblast that penetrates the inner
myometrium of the uterus (Carter and Pijnenborg, 2011). It is
hypothesized that this feature is related to the increased demand for
resources by the developing fetus to feed the expanded size of the
human brain (Carter and Pijnenborg, 2011; Robillard et al., 2002).
Humans have the largest brain in terms of relative size and this
differs markedly from closely related great apes (Boddy et al., 2012).
Deep penetration of the myometrium occurs in chimpanzees and gorillas,
but this penetration may be accelerated in humans (Carter and
Pijnenborg, 2011).

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
preeclampsia is a human disease that occurs in 4\% of pregnancies in
the United States. Preeclampsia can be attributed to the lack of
adequate invasion of maternal tissue, and is noted by the onset of
hypertension and proteinuria (citation). Some cases of preeclampsia
can develop into eclampsia, a more severe condition resulting in
seizures during pregnancy and death. Lack of invasion in preeclampsia
may be attributed to the inefficiency of growth regulators as well as
the inability to sufficiently suppress the maternal immune response.
Although there are speculative cases of eclampsia in great apes
(Baird, 1981; Stout and Lemmon, 1969; Thornton and Onwude, 1992), it
is exceedingly rare in these species (Carter and Pijnenborg, 2011). It
is possible that a comparative evolutionary approach to gene
expression can identify genes that have divergent expression patterns
from other mammals as well as association with preeclampsia.

>>>>

One symptom
of this maternal-fetal conflict is its potential affect on human
health \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. 
In addition to having an invasive
hemochorial placenta, humans possess a deeply invasive trophoblast
layer that can penetrate the myometrium of the uterus 
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
 Defects in placental growth are known to be
associated with the devastating condition of preeclampsia, which can
cause hypertension, proteinuria, and has the potential to lead to
death of the mother and/or infant \DLA{(citation)}. Although cases have been
documented in chimpanzees and gorillas \DLA{(citation)}, preeclampsia has a
far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States \DLA{(citation)}.
Chimpanzees and gorillas also have deeply invasive trophoblasts,
although the progression of invasion in great apes may be slower than
humans \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however, large-scale expression
studies of great ape placentas have not been conducted. The increasing
affordability of next-generation sequencing technology has enabled the
large-scale study of specific tissues from a wide array of species to
address broader phylogenetic questions in the evolution of gene
expression \cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use Illumina RNA-Seq
technology to sequence and quantify the mRNA environment of six
mammalian placentas at term, including that of the human, bonobo (the
sister species to the chimpanzee), dog, cow, African elephant, and
opossum, representing marsupials as well as three major branches of
eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation \DLA{(Table 1)}. We use this data, along with existing RNA-Seq
data from mouse \cite{Wang.ea2011:surveyfornovelimprinted}, to address the major differences
between expression in marsupial and eutherian mammals at term,
identify the evolution of expression in important placental genes such
as hormones and galectins, identify novel lineage-specific expression
patterns, and quantify human specific differences in gene expression
that may be related to obstetrical disorders such as preeclampsia.

@@

Fetal tissue was collected from the placenta of six mammalian species
(\textit{Homo sapiens} [human], \textit{Pan paniscus} [bonobo], \textit{Canis familiaris}
[domestic dog], \textit{Bos taurus} [domestic cow], \textit{Loxodonta africana} [African
savanna elephant], \textit{Monodelphis domestica} [gray short-tailed
opossum], \textit{Nannospalax gallili} [?], \textit{Spalax carmelli} [?]).
Information on specimen ID, place and time of collection,
and collector is shown for each sample in \DLA{Supplemental Table 1}. In the
case of \textit{Pan}, \textit{Bos}, and \textit{Loxodonta}, the placenta was collected at term
upon expulsion from the mother following delivery. Placental tissue
from \textit{Homo sapiens} was collected upon delivery via a caesarian section.
Near-term placental tissue from \textit{Canis familiaris} was collected upon
the death of the mother from a car accident. Placental tissue from
\textit{Monodelphis domestica} was sampled at fetal stage 32, roughly one day
prior to birth. Only the fetal portion of the placenta was sampled at
the maternal-fetal interface (i.e. villous portion for \textit{Homo}, \textit{Pan},
\textit{Canis}, and \textit{Loxodonta}; cotyledon for \textit{Bos}; diffuse placenta for
\textit{Monodelphis}). All tissues were collected and stored in RNAlater (Life
Technologies, Carlsbad, CA) until RNA isolation could be performed.
Tissues were homogenized in Trizol, and total RNA was isolated
following the manufacturer’s instructions. RNA was further purified
using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction with the
RNase-Free DNase Set (Qiagen, Valencia, CA), following the
manufacturer’s protocol to remove any contaminating DNA from the
sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
Clara, CA) to confirm concentration and quality of the isolation. RNA
quality was assessed using A260/A280 and A260/A230 absorbance values.

\DLA{Add details for new sequences and fix numbers}

@@

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and standard Illumina protocol. Paired-end sequencing was
performed using the Illumina Genome Analyzer II with an insert size of
230 base pairs (bp). \textit{Bos}, \textit{Loxodonta}, and
\textit{Monodelphis} were sequenced with read lengths of 75 bp;
\textit{Canis}, \textit{Homo}, and \textit{Pan} were sequenced with
read lengths of 150 bp.

@@

\DLA{This isn't how these were analyzed anymore; fix this.}

The quality of reads was evaluated using FASTQC v. 0.9.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}). Following
this, reads were trimmed based on quality in Galaxy using the FASTQ
Trimmer v. 1.0.0 (Blankenberg et al., 2010). All transcriptomes were
then aligned to reference genomes and assembled using the Tuxedo suite
of next-generation sequencing analysis tools. Annotated reference
genomes were downloaded from Ensembl v. 64 (Flicek et al., 2012),
including Homo sapiens GRCh37, Pan troglodytes CHIMP2.1 (6X coverage),
Canis familiaris BROADD2 (7.6X), Bos taurus UMD3.1 (9.5X), Loxodonta
africana loxAfr3 (7X), and Monodelphis domestica BROADO5 (6.8X). The
genome of the closely-related Pan troglodytes was used to map the
transcriptome of Pan paniscus; the separation between the two
recognized Pan species is generally perceived to be ~1 Mya (Prufer et
al., 2012). Reads were aligned to reference genomes using Bowtie v.
0.12.7 (Langmead et al., 2009) and TopHat v. 1.3.3 (Trapnell et al.,
2009). Cufflinks v. 1.2.1 was then used to assemble transcripts and
quantify expression values using fragments per kilobase of exon per
million mapped reads (FPKM) (Trapnell et al., 2010). RNA-Seq metrics
were computed using Picard (http://picard.sourceforge.net/). For
comparison, we also downloaded the placenta Illumina RNA-Seq
transcriptome of a mouse from (Wang et al., 2011). Raw reads were
mapped as above using the Ensembl v. 64 annotation of the Mus musculus
(house mouse) genome (NCBI m37).

@@

The human placenta is distinctive from many other primates in that it
possesses a deeply invasive trophoblast that penetrates the inner
myometrium of the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
hypothesized that this feature is related to the increased demand for
resources by the developing fetus to feed the expanded size of the
human brain (Carter and Pijnenborg, 2011; Robillard et al., 2002).
Humans have the largest brain in terms of relative size and this
differs markedly from closely related great apes (Boddy et al., 2012).
Deep penetration of the myometrium occurs in chimpanzees and gorillas,
but this penetration may be accelerated in humans (Carter and
Pijnenborg, 2011).

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
preeclampsia is a human disease that occurs in 4\% of pregnancies in
the United States. Preeclampsia can be attributed to the lack of
adequate invasion of maternal tissue, and is noted by the onset of
hypertension and proteinuria (citation). Some cases of preeclampsia
can develop into eclampsia, a more severe condition resulting in
seizures during pregnancy and death. Lack of invasion in preeclampsia
may be attributed to the inefficiency of growth regulators as well as
the inability to sufficiently suppress the maternal immune response.
Although there are speculative cases of eclampsia in great apes
(Baird, 1981; Stout and Lemmon, 1969; Thornton and Onwude, 1992), it
is exceedingly rare in these species \cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It
is possible that a comparative evolutionary approach to gene
expression can identify genes that have divergent expression patterns
from other mammals as well as association with preeclampsia.
